\chapter{Introduction and Objectives (10 marks)}%
\label{chp:introduction}
The Vehicle Routing Problem (VRP) is a classical combinatorial optimization problem 
concerned with determining the most efficient set of routes for a fleet of vehicles 
to service a group of customers. The primary objective is typically to minimize total 
routing cost, although other factors, such as fuel consumption or fleet size, are often 
considered in practical applications.

The Dynamic Constrained Vehicle Routing Problem with Time Windows (DCVRPTW) is a more complex
variant of VRP that incorporates dynamic elements, such as customers revealed in real time,
as well as operational constraints, such as customer service time windows. This problem is
particularly relevant in modern logistics and transportation, where companies must continuously
adapt to changing conditions while meeting increasingly strict delivery expectations.

This study investigates whether AI-based approaches can effectively support decision-making in
highly dynamic scenarios with tight time windows.

To this end, we compare three approaches: a purely heuristic method, a purely AI-based method,
and a hybrid method in which AI-generated solutions are used as a warm start for subsequent
heuristic refinement. Performance is evaluated using three metrics: responsiveness (time taken
to service newly revealed customers), lateness (both the frequency and magnitude of delays), and
rejection rate (the proportion of customers that cannot be feasibly served).

Experiments are conducted on the Solomon dataset, a widely used benchmark for VRPTW problems.
Beyond its academic relevance, this project is also practically significant for logistics and
transportation, as it explores the potential of AI to improve operational efficiency in dynamic
routing environments.